---
Reproducible Research: Peer Assessment 1

Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit , Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain underutilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.
---


## Loading and preprocessing the data

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data is a comma delimited file, it includes 17,568 observations of 3 variables:

steps: Number of steps taken in a 5 min interval
date: The date when the measurement was taken in YYY-MM-DD format
interval: Identifier for the 5-min interval in which the measurement was taken
First, we read the date coercing the date column to character rather than factor


```{r, cache = TRUE}
activityData <- read.csv ("activity.csv", header = T, sep = ",", stringsAsFactors = F)

```

Now we convert the date column to the appropriate format:

```{r, cache = TRUE}
activityData$date <- as.Date(activityData$date, "%Y-%m-%d")
str(activityData)
```

## What is mean total number of steps taken per day?

```{r, cache = TRUE}
library (dplyr)
AvgDay <- activityData %>% group_by(date) %>%
          summarize(total.steps = sum(steps, na.rm = T), 
                  mean.steps = mean(steps, na.rm = T))
```                  
Once the summaries are calculated, we can construct the histogram of the total steps:

```{r, cache = TRUE}
library(ggplot2)
g <- ggplot(AvgDay, aes(x=total.steps))
g + geom_histogram(binwidth = 2500) + theme(axis.text = element_text(size = 12),  
      axis.title = element_text(size = 14)) + labs(y = "Frequency") + labs(x = "Total steps/day")
``` 

```{r, cache = TRUE}
summary(AvgDay$total.steps)
``` 

## What is the average daily activity pattern?

We group the data by interval this time and then calculate the mean of each interval goup:

```{r, cache = TRUE}
AvgInterval <- activityData %>% group_by(interval) %>%
      summarize(mean.steps = mean(steps, na.rm = T))
g <- ggplot(AvgInterval, aes(x = interval, y = mean.steps))
g + geom_line() + theme(axis.text = element_text(size = 12), 
      axis.title = element_text(size = 14, face = "bold")) + 
      labs(y = "Mean number of steps") + labs(x = "Interval")
``` 
We can observe the largest amount of steps occurs between time intervals 500 and 1000. The maximum average number of steps is: 206 and occurs in time interval #835

## Imputing missing values

We noticed that there are missing values when we printed the first few rows of the activityData variable, but so far we have not determined how many values are missing. The following lines will calculate the percentage of missing data as well as the number of rows that contain an NA.

```{r, cache = TRUE}
mean(is.na(activityData$steps))
``` 

```{r, cache = TRUE}
sum(is.na(activityData$steps))
``` 

In order to fill in missing values we check at each row if the column interval is NA, when the condition is true we look for the corresponding interval (index), we search for this particular interval in the AvgInterval data and extract it to a temporary variable values. Last we choose only the column of interest from values, which is the mean.steps and assign this number to the corresponding position in the newData set. We use a for loop to run through all the rows. (Note: there may be a more elegant way to do this perhaps using apply but couldnt make it work)

```{r, cache = TRUE}
for (i in 1:nrow(newData)) {
      if (is.na(newData$steps[i])) {
            index <- newData$interval[i]
            value <- subset(AvgInterval, interval==index)
            newData$steps[i] <- value$mean.steps
      }
}
head(newData)
```

## Are there differences in activity patterns between weekdays and weekends?

Different weekend vs weekday patterns are expected as people, in general, have a different set of activities on weekends.
In order to find the specific patterns for each set of days, we will identify the weekdays from the weekend data. First, we create a new column in newData containing the values weekend or weekday:
```{r, cache = TRUE}
newData$day <- ifelse(weekdays(newData$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
```
Next we create two subsets, one containing the weekend and one containing the weekday data:

```{r, cache = TRUE}
wkend <- filter(newData, day == "weekend")
wkday <- filter(newData, day == "weekday")
```

Then, similarly to section 2, we group by the intervals and calculate the mean number of steps for each time interval. Since the day column is lots during the grouping, we add it again to the wkend and wday dataframes. Lastly, we merge both data sets into one named newInterval
```{r, cache = TRUE}
wkend <- wkend %>%
      group_by(interval) %>%
      summarize(mean.steps = mean(steps)) 
wkend$day <- "weekend"

wkday <- wkday %>%
      group_by(interval) %>%
      summarize(mean.steps = mean(steps)) 
wkday$day <- "weekday"

newInterval <- rbind(wkend, wkday)
newInterval$day <- as.factor(newInterval$day)
newInterval$day <- relevel(newInterval$day, "weekend")
```

The two panel plot is now created, using the day column as a factor to spearate the weekday from the weekend timeseries.
```{r, cache = TRUE}
g <- ggplot (newInterval, aes (interval, mean.steps))
g + geom_line() + facet_grid (day~.) + theme(axis.text = element_text(size = 12), 
      axis.title = element_text(size = 14)) + labs(y = "Number of Steps") + labs(x = "Interval")
```

We observe that, as expected, the activity profiles between weekdays and weekends greatly differ. During the weekdays, activity peaks in the morning between 7 and 9 and then the activity remains below ~100 steps. In contrast, the weekend data does not show a period with particularly high level of activity, but the activity remains higher than the weekday activity at most times and in several instances it surpases the 100 steps mark and it is overall more evenly distributed throughout the day.